Polka-dotted Stars: a Hierarchical Model for Mapping Stellar Surfaces Using Occultation Light Curves and the Case of TOI-3884

### 2025-03-05 src/tex/ms.tex
first draft

% Author list
\author[0000-0002-6650-3829]{Sabina Sagynbayeva}
\email{sabina.sagynbayeva@stonybrook.edu}
\affiliation{\sbu}
\affiliation{\cca}

@@

% Abstract with filler text
\begin{abstract}

\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}
The quest to unravel the mysteries of distant planetary systems has led to a fast-evolving field of research, where exoplanet studies are at the forefront of 
astronomical investigations. The detection of atmospheres on Earth analogs using transmission spectroscopy is a challenging task due to the confounding 
effects of stellar contamination (see, e.g., \cite{Ducrot2018, Morris2018}). Stellar magnetic activity, particularly in the form of starspots, can significantly 
influence the observed transmission spectra, making it difficult to disentangle planetary atmospheric signals from those originating from the host star. 
Starspots exhibit variations in both time and wavelength, mainly due to stellar rotation. As a result, the identification of starspots has the potential to 
complicate the analysis and interpretation of exoplanet observations.

Therefore, characterizing the behavior of stellar surface features that arise due to magnetic activity like starspots and flares is instrumental for improving 
planet detection efficiency in photometric, spectral, and radial velocity studies. Understanding starspots is fundamental and imminent for the ongoing 
space mission JWST, which will characterize tens to hundreds of exoplanet atmospheres around stars with strong variability. In fact, recent JWST 
results (\cite{Lim2023}) on the atmosphere of TRAPPIST-1b showed that accurate measurements have so far been substantially stymied by stellar contamination, 
and we need more accurate models to account for it. To fully capitalize on the unprecedented capabilities of JWST and accurately interpret the obtained 
transmission spectra, it is crucial to develop a comprehensive understanding of stellar magnetic activity and its impact on the observed signals.

Moreover, characterizing the surfaces of exoplanet-host stars is crucial for understanding the environments and evolution of their planetary 
systems. For example, the Sun's magnetic field drives the star's activity cycles, the heating of its outer atmosphere and the solar wind,
which affects the Earth's magnetosphere and biosphere \citep{Babcock1961,Charbonneau2014}. 
Therefore, constructing a catalogue of the magnetic properties of other stars similar to the Sun would support the quest towards understanding 
the population of detected exoplanets and the development of a highly generalizable theory on stellar magnetic dynamo.

The reverse statement is also true. The exisiting popultion of exoplanets along with their orbital parameters can help characterize their host-stars
in terms of their surface magnetic activity. Stellar surface features like spots and flares impact observables in planet-detection research including photometric 
variability, spectral line profiles, and radial velocity shifts. However, directly resolving stellar surfaces remains challenging. 
While high-resolution spectroscopy provides the most robust method of determining stellar parameters, obtaining such data for large samples remains 
observationally expensive, but photometric observations can come to rescue. Some recent works offered invaluable insights into the topographical 
characteristics of stars and their magnetic activity, responsible for the observed variatians on their surfaces. For example, the influence of starspots 
can be seen through photometric data acquired from Kepler or TESS. \cite{Luger2021b} investigated the star spots features on the stellar surfaces using 
Gaussian Process (GPs) with a \emph{physically interpretable} kernel. They argue that the complex degeneracies make it impossible to determine the 
precise configuration of starspots and other features on a star's surface solely based on its rotational light curve, but we don't have to be 
concerned with the specific characteristics of an individual spot, as such information would have limited practical utility. Instead, we often consider 
(either explicitly or implicitly) the properties of a star spot as a sample drawn from an underlying parent distribution that governs aspects like the average 
and dispersion of spot radii. The parameters that control this distribution are the ones that can be predicted through physical principles, 
making them the primary focus of our interest.

\cite{Luger2021b} also showed that it is hard to put constrains on starspot properties if you only have a light curve for 
one star, it is rather usefule though to perform the same analysis for an ensemble of stars. However, photometric light curves capturing planetary transits encode 
a wealth of information about the stellar surface and the starspots present on it. Transits, during which the planet occults different regions of 
the stellar disk, provide a unique opportunity to probe the physical characteristics of starspots. Exoplanet host stars are \textit{uniquely} suited 
for photometric studies of starspot properties: spot features of observed exoplanet host stars in transit can provide valuable 
constraints on spot properties (see Figure \ref{fig:cartoon}) -- brightness, size, and latitude -- while fitting a single light curve for a star 
with no observed transit introduces too many degeneracies.

In this work, we utilize two powerful computational tools, \texttt{starry} and \texttt{StarryProcess}, to model the light curves of spotted stars and 
their transiting exoplanets. We then extend these models to create a comprehensive framework that combines both stellar variability and planetary transits.

The \texttt{starry} code, developed by \cite{Luger2019}, provides an efficient method for computing light curves of stars and planets. 
It uses a spherical harmonic framework to represent the surface brightness distribution of celestial bodies, allowing for fast and accurate 
calculations of rotational light curves and occultations. \texttt{starry} is particularly useful for modeling the effects of stellar spots 
and planetary transits, as it can handle complex surface brightness patterns and compute their resulting light curves with high precision.

Building upon \texttt{starry}, \cite{Luger2021b} introduced \texttt{StarryProcess}, a probabilistic model for stellar variability. \texttt{StarryProcess} 
treats the distribution of spots on a stellar surface as a Gaussian process in spherical harmonic space. This approach allows for the efficient 
modeling of stellar variability over long time scales, accounting for the evolution and rotational modulation of starspots. \texttt{StarryProcess} 
provides a flexible framework for inferring stellar and starspot properties.

Our model combines the capabilities of both \texttt{starry} and \texttt{StarryProcess}, integrating them with a transit model to 
create a comprehensive framework for analyzing light curves of spotted stars with transiting exoplanets. By leveraging the spherical harmonic 
representation from \texttt{starry} and the probabilistic approach of \texttt{StarryProcess}, we can simultaneously model the stellar variability, 
spot distribution, and planetary transits. This combined approach allows us to account for the effects of stellar spots on transit light curves, 
including spot-crossing events, while also constraining the overall stellar variability pattern. This integrated model enables us to derive more 
accurate and precise constraints on both stellar and planetary parameters. It accounts for the interplay between stellar activity and planetary transits, 
allowing us to disentangle the effects of spots from true transit features. Furthermore, by modeling multiple transits simultaneously with the 
underlying stellar variability, we can better constrain parameters such as stellar inclination, obliquity, and spot distribution, 
while also improving our estimates of planetary properties.

\begin{figure*}[hbt!]
    \includegraphics[width=\textwidth]{figures/star-planet-cartoon.png}
        \caption{Light curves from the kinds of stars shown on the left panel (these light curves were calculated analytically 
        using \texttt{starry} \citep{Luger2019}): (a) the light curve shows a flat flux if the star has no surface features and no planet orbiting it; 
        (b) a transit in the light curve of the star that hosts an exoplanet; (c) if the star has one orbiting planet and one starspot under the planet's 
        trajectory then it is seen as a bump in a transit signal; (d) the light curve becomes complicated when the star has multiple spots.}
        \label{fig:cartoon}
    \end{figure*}

The light curves obtained from telescopic observations often provide insufficient information for comprehensive photometric studies of stellar 
magnetic fields, yet they also yield an overabundance of data for simple forward modeling. To address the first issue, we focus on capturing the essence 
of the variability caused by the magnetic dynamo rather than solving it in great detail (as per the argument in \cite{Luger2021b}). The second challenge can 
be overcome by marginalizing over (integrating out) the properties of each individual spot on a star to infer the characteristics of the overall spot distribution. 
This could be accomplished by modeling the properties of each spot and generating 
the corresponding light curves with a model like \texttt{starry}, which does exactly that by constructing a disk-integrating signal 
with spherical harmonics. Then one can use a posterior sampling scheme, such as Markov Chain Monte Carlo (MCMC), 
while incorporating hyperparameters that control the distribution of spot properties across the entire population. The marginal posteriors for 
the hyperparameters would ultimately provide the information we seek. Bayesian modeling provides a principled framework for incorporating prior 
knowledge and uncertainties into the inference process, allowing us to robustly estimate the hyperparameters governing the distribution of star spot 
properties while naturally accounting for the complex degeneracies and multi-modality inherent in the problem.
However, in practice, the degeneracies and often extreme multi-modality of the distributions of individual spot properties would render this approach 
quite challenging and computationally expensive. That's why we harness the elegant machinery like Gaussian processes (GPs) to perform 
this marginalization on our behalf, simplifying the process and reducing the computational burden.

In Figure \ref{fig:nullspace}, we present an analysis similar to Figure 4 in \cite{Luger2021b}, but with the addition of a planetary transit. 
This figure illustrates the decomposition of a stellar surface map with a transiting planet into its preimage and null space components, 
along with their corresponding contributions to the light curve. Crucially, we observe that the planetary trajectory covers the visible part in the preimage 
component but absent in the null space. The transit path effectively "carves out" a strip on the stellar surface in nullspace but that is 
fully contained within the preimage.

\begin{figure*}[hbt!]
    \includegraphics[width=\textwidth]{figures/nullspace-preimage.png}
        \caption{The breakdown of a surface map into its constituent components: the original map (left), its preimage (center), and null space (right). 
        Accompanying each set is the corresponding impact on the star's rotational light curve. The preimage represents surface patterns that 
        directly influence the observed light curve, while the null space encompasses patterns that have no effect on the flux measurements. 
        It's noteworthy that a significant portion of the surface features fall within the null space, meaning they do not contribute to the light curve 
        that we can observe and measure. The blck line shows the planetary trajectory, so what is important is that the preimage contains the full
        information under planetary trajectory, while the nullspace is missing it.}
        \label{fig:nullspace}
    \end{figure*}
% In this work, we develop a Gaussian process framework to model the surfaces of exoplanet-host stars based on time-series photometry. We utilize 
% expressive kernel function from \cite{Luger2021b}. We first use this model on a simulated data to test the model, then we apply the model to an active K-dwarf
% HAT-P-11.

% The planet-hosting star HAT-P-11 has been extensively studied for its star spots. HAT-P-11 is orbited by a Neptune-sized exoplanet, HAT-P-11b, 
% which was discovered in 2008 via the transit method \citep{Bakos2010,Deming2011}. The transiting nature of HAT-P-11b has allowed for detailed 
% characterization of the star spot activity on HAT-P-11, as star spots that are occulted during the planet's transit cause detectable brightening in 
% the transit light curve \citep{Morris2017}.

%
% \section{The Data}
% The data is collected by the \emph{Kepler} mission and we extracted it using
% \texttt{lightkurve}, a Python package for Kepler and TESS data analysis \citep{lightkurve}.
% %
% \begin{figure}[ht!]
%     \script{TransitFitsWithStarry.py}
%     \begin{centering}
%         \includegraphics[width=\linewidth]{figures/TransitFitsWithStarry.pdf}
%         \caption{
%             Transit fits with starry -- no GPs.
%         }
%         \label{fig:TransitFitsStarry}
%     \end{centering}
% \end{figure}
%
\section{Hierarchical Bayesian model}
\label{sec:model}
In this section, we describe our Gaussian Process (GP) model and the likelihood calculation process used to estimate the model parameters.
This essentially means that we approximate the likelihood 
as a multidimensional Gaussian distribution characterized by a mean $\pmb{\mu}$ and covariance $\pmb{B}$ (for a comprehensive 
derivation of the Gaussian Process, refer to \cite{Luger2021b}):
%
\begin{align}
    \mathcal{L}\left(i_\star, P_\star, \mathbf{u}, \pmb{\theta}_\bullet\right) \sim
    \mathcal{N}\Big(
    \pmb{\mu}\left(i_\star, P_\star, \mathbf{u}, \pmb{\theta}_\bullet\right),
    \,
    \pmb{B}\left(i_\star, P_\star, \mathbf{u}, \pmb{\theta}_\bullet\right)
    \Big).
\end{align}
%
To infer the physical properties of the starposts, we want to obtain an underlying distribution of the physics conditioned on the provided data, 
which is known as the posterior distribution of the model.
% \subsection{Model for non-evolvng surfaces} \label{sec:nonevolmodel} 
We want to solve for a large set of parameters that includes the GP hyperparameters, the star's and the planet's orbital parameters, respectively. 
\begin{linenomath}\begin{align}
    \label{eq:largetheta}
    \pmb{\Theta}
     & =
    \left(
    \theta_\bullet
    \,\,\,
    \theta_\star
    \,\,\,
    \theta_p
    \right)^\top
    \quad,
\end{align}\end{linenomath}

@@

We represent the GP hyperparameters as \emph{physically interesting} set of parameters $\pmb{\theta}_\bullet$ \citep{Luger2021b}:
%
\begin{linenomath}\begin{align}
        \label{eq:thetaspot}
        \pmb{\theta}_\bullet
         & =
        \left(
        \mathbb{n}
        \,\,\,
        \mathbb{c}
        \,\,\,
        \mathbb{a}
        \,\,\,
        \mathbb{b}
        \,\,\,
        \mathbb{r}
        \right)^\top
        \quad,
    \end{align}\end{linenomath}
%
where $\mathbb{n}$ is the number of starspots, $\mathbb{c}$ is their contrast (defined as the intensity difference between the spot and the 
background intensity, as a fraction of the background intensity), and $\mathbb{r}$ is the radius
of the spots. $\mathbb{a}$ and $\mathbb{b}$ are the normalized parameters of the Beta distribution in $\cos\phi$, which is the probability density function (PDF) 
for the latitude $\phi$ of the spots. These parameters have a one-to-one correspondence to the mode $\mu_\phi$ and standard deviation $\sigma_\phi$ of the 
distribution in $\phi$, allowing for a concise representation of the latitudinal distribution characteristics.
The Beta distribution in $\cos\phi$ which has hyperparameters $\alpha$ and $\beta$, and the PDF given by
%
\begin{align}
    \label{eq:cosphi-pdf}
    p \big(\cos\phi \, \big| \, \alpha, \beta \big)
     & =
    \dfrac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)}
    (\cos\phi)^{\alpha - 1}
    (1 - \cos\phi)^{\beta - 1}
    \quad,
\end{align}
%
where $\Gamma$ is the Gamma function. The $\alpha$ and $\beta$ are derived from the normalized parameters 
%
\begin{align}
    \label{eq:beta2gauss}
    \alpha & = \exp\left({K_{00} + (\ln\frac{1}{2})\mathbb{a}}\right)
    \nonumber                                                 \\
    \beta  & = \exp\left({\ln\frac{1}{2} + (10 - \ln\frac{1}{2})\mathbb{b}}\right)
    \quad,
\end{align}
%
with inverse transform
%
\begin{align}
    \label{eq:gauss2beta}
    \mathbb{a} & \equiv \frac{\ln\alpha}{\ln\frac{1}{2}}
    \nonumber                                             \\[0.5em]
    \mathbb{b} & \equiv \frac{\ln\beta - \ln\frac{1}{2}}{10 -\ln\frac{1}{2}}
\end{align}
%
The parameters $\mathbb{a}$ and $\mathbb{b}$ are both constrained to values between 0 and 1, making them convenient for sampling during the inference process. 
However, $\mathbb{a}$ and $\mathbb{b}$ do not have a straightforward relationship with physically meaningful quantities. In many situations, it is preferable to parametrize 
the latitude distribution using two parameters: $\mu_\phi$, which controls the central latitude, and $\sigma_\phi$, which governs the dispersion of the 
spots' latitudes. Here, both the mean $\mu_\phi$ and variance $\sigma^2_\phi$ can be derived from the Beta distribution as
%
\begin{align}
    \label{eq:mean_beta}
    \mu_\phi
     & =
    \dfrac{\alpha}{\alpha+\beta}
    \quad,
\end{align}
%
\begin{align}
    \label{eq:var_beta}
    \sigma^2_\phi
     & =
    \dfrac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}
    \quad,
\end{align}
%

@@

For faster calculations we rewrite the likelihood function using \emph{the matrix conversion lemma}, also known as 
Woodbury-Sherman-Morrison identity (see, e.g., \cite{Hogg2020}), which gives an expression for the inverse $\pmb{B}^{-1}$ of the marginalized likelihood variance:
%
\begin{linenomath}\begin{align}
    \label{eq:Hoggtrick}
    \pmb{B}^{-1} = \pmb{C}^{-1} - \pmb{C}^{-1} \pmb{M} (\pmb{\mathbb{I}} 
    & + \pmb{\Lambda} \pmb{M}^T \pmb{C}^{-1}\pmb{M})^{-1}\pmb{\Lambda} \pmb{M}^T\pmb{C}^{-1}
\end{align}\end{linenomath}
%

@@

\subsection{Step-by-step model instruction}
In this section, we provide a comprehensive, step-by-step description of our model's methodology. 
This process outlines the key stages in our analysis, from initial data preparation to the final likelihood computation:

\begin{enumerate}
    \item Acquire the target system's light curve data.
    \item Preprocess the data by binning the out-of-transit observations to optimize computational efficiency while preserving signal fidelity.
    \item Construct a design matrix incorporating both planetary (transit) and stellar orbital parameters, encapsulating the geometry of the star-planet system.
    \item Establish physically motivated hyperparameters and prior distributions based on current astrophysical understanding and observational constraints.
    \item Implement a sampling algorithm to draw from the defined prior distributions, exploring the parameter space efficiently.
    \item Generate synthetic light curves from the sampled parameters, simulating the observed stellar and planetary signals.
    \item Compute the covariance matrix for the Gaussian process, which models the stochastic components of stellar variability.
    \item Assemble the inverse of the full covariance matrix, combining deterministic and stochastic elements of the model based on Equation \ref{eq:Hoggtrick}.
    \item Calculate the likelihood of the observed data given the model parameters, quantifying the agreement between our model and the observations with Equation \ref{eq:log-likeRodrigo}.
\end{enumerate}

@@

\section{Results: synthetic datasets}
\label{sec:results-synthetic}
\subsection{Priors}
Before presenting the results, it is essential to elucidate the prior distributions employed throughout this work. 
We have implemented multiple priors in this work, carefully selected to maintain a balance between informativeness and generalizability. 
These priors serve to encapsulate our existing knowledge and assumptions about the parameters of interest, while allowing for sufficient flexibility 
to avoid computational cost.
\begin{itemize}
    \item $\sim\mathcal{U}(low, high)$: a uniform distribution.
    \item \textit{Planetary Inclination}: Considering that $i_p=\arccos{b}$, where $b$ is the impact parameter,
    the prior on the inclination is defined via the uniform prior $\sim\mathcal{U}(-b_{max}, -b_{max})$ on $b$, where $b_{max} = R_{\star} / a$.
    \item \textit{Period}: a prior on $P$ and $P_{\star}$. Instead of directly constraining the period, this prior works in log-space to ensure the period 
    remains positive and to make the distribution more symmetric. The actual prior is a uniform distribution in log-space, bounded by the log of the true 
    value plus or minus some fraction. The final period is determined by exponentiating the result of this log-uniform distribution, 
    which effectively creates a kind of log-uniform prior on the period itself.
    \item \textit{Logarithmic}: a log-uniform distribution.
    \item \textit{Stellar Angle}: a normal distribution on orientation vectors $\tilde{x}$ and $\tilde{y}$ and a half-normal distribution on $\tilde{z}$. See Section
    \ref{sec:obl-inc} for the full explanation.
\end{itemize}

@@

We utilized the \texttt{Python}-based implementation of the GP introduced by \cite{Luger2021b} to perform inference on the synthetic dataset. 
We used No-U-Turn Sampling, a variant of Hamiltonian Monte Carlo 
(NUTS; \cite{Duane1987,Hoffman2011}) to do posterior inference on our synthetic dataset.

>>>>

% Author list
\author[0000-0002-6650-3829]{Sabina Sagynbayeva}
\email{sabina.sagynbayeva@stonybrook.edu}
\affiliation{\sbu}
\affiliation{\cca}
\affiliation{Kavli Institute for Theoretical Physics, University of California, Santa Barbara, CA 93106, USA}

@@

% Abstract with filler text
\begin{abstract}
    We present \texttt{StarryStarryProcess}, a novel hierarchical Bayesian framework for mapping stellar surfaces using exoplanet transit light curves. 
    While previous methods relied solely on stellar rotational light curves -— which contain limited information about spot properties -- our approach leverages 
    planetary transits as probes of stellar surfaces. When a planet crosses a spot during transit, it creates a distinctive bump in the light curve that 
    directly reveals spot properties. Our model integrates planetary transit modeling with stellar variability analysis by combining 
    the spherical harmonic representation from \texttt{starry}, the probabilistic approach of \texttt{StarryProcess}, and a comprehensive 
    transit model that accounts for spot-crossing events. We demonstrate through synthetic data experiments that our model successfully recovers spot distributions, 
    stellar orientation, and spot physical properties. We extend the framework to handle evolving stellar surfaces through time-dependent modeling. 
    Applying our method to TESS observations of TOI-3884, we find evidence for high-latitude spot concentrations and significant spin-orbit misalignment. The transit-based 
    approach overcomes fundamental limitations of previous models by providing constraints on spot properties that would remain hidden in the null space of 
    rotational light curves alone. This methodology enables more accurate exoplanet characterization by disentangling stellar activity from planetary signals 
    while simultaneously providing insights into stellar magnetic activity patterns.
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}
The detection of atmospheres on Earth analogs using transmission spectroscopy is challenging due to stellar 
contamination effects \citep{Ducrot2018, Morris2018}. Stellar magnetic activity, particularly starspots, significantly 
influences observed transmission spectra, making it difficult to distinguish planetary atmospheric signals from stellar ones. 
Starspots exhibit variations in both time and wavelength primarily due to stellar rotation, potentially complicating the analysis 
and interpretation of exoplanet observations.

Therefore, characterizing stellar surface features arising from magnetic activity—-such as starspots and flares—-is crucial for 
improving planet detection efficiency across photometric, spectral, and radial velocity studies. Understanding starspots is 
especially important for the JWST mission, which aims to characterize numerous exoplanet atmospheres around stars with strong 
variability. Recent JWST results on TRAPPIST-1b \citep{Lim2023} demonstrated that accurate measurements have been substantially 
hindered by stellar contamination, highlighting the need for more precise models to account for these effects. To fully leverage 
JWST's capabilities and accurately interpret transmission spectra, we must develop a comprehensive understanding of stellar 
magnetic activity and its impact on observed signals.

Recent works have provided valuable insights into the topographical characteristics of stars and their magnetic activity. 
For example, starspot influence can be observed through photometric data from Kepler or TESS. The \texttt{starry} code 
\citep{Luger2019} offers an efficient method for computing light curves of stars and planets using a spherical harmonic 
framework to represent surface brightness distribution, enabling fast and accurate calculations of rotational light curves 
and occultations. Building upon \texttt{starry}, \citet{Luger2021b} introduced \texttt{StarryProcess}, a probabilistic model 
for stellar variability that treats spot distribution as a Gaussian process in spherical harmonic space, allowing efficient 
modeling of stellar variability over extended time scales.

With \texttt{StarryProcess}, \citet{Luger2021b} investigated stellar spot features using Gaussian Processes with 
a physically interpretable kernel. They argued that complex degeneracies make it impossible to determine the precise 
configuration of starspots solely from rotational light curves. Instead, they suggested focusing on the statistical properties 
of spot distributions rather than individual spot characteristics. However, they also demonstrated that constraining starspot 
properties from a single star's light curve is difficult; an ensemble analysis of multiple stars is more informative.

While previous models like \texttt{StarryProcess} relied solely on stellar rotational light curves—which contain limited 
information about spot properties—our approach introduces a critical advancement: we leverage planetary transits as probes of 
stellar surfaces. Exoplanet host stars are \textit{uniquely suited} for photometric studies of starspot properties because:
(1) Planetary transits effectively scan different regions of the stellar disk;
(2) Transit light curves contain encoded information about the stellar surface and starspots;
(3) Spot-crossing events during transits provide direct constraints on spot brightness, size, and latitude.

This transit-based approach overcomes the fundamental limitation of previous models that analyzed stellar light curves alone, 
where too many degeneracies exist to precisely characterize spots. As illustrated in Figure \ref{fig:cartoon}, when a planet 
crosses a spot during transit, it creates a distinctive bump in the light curve that directly reveals spot properties.

In this work, we develop \texttt{StarryStarryProcess}, a novel framework that integrates planetary transit modeling with 
stellar variability analysis. Our model combines:
\begin{itemize}
    \item The spherical harmonic representation from \texttt{starry}
    \item The probabilistic approach of \texttt{StarryProcess}
    \item A comprehensive transit model that accounts for spot-crossing events
\end{itemize}

By simultaneously modeling stellar variability, spot distributions, and planetary transits, our approach disentangles 
spot effects from transit features, resulting in more accurate constraints on both stellar and planetary parameters. 
When modeling multiple transits alongside underlying stellar variability, we can better determine stellar inclination, 
obliquity, and spot distribution patterns.

Figure \ref{fig:nullspace} illustrates a key insight of our approach: while previous methods struggled with the null 
space problem (where many surface features don't affect the observable light curve), planetary transits effectively 
"carve out" paths on the stellar surface that reveal information otherwise hidden in the null space. The transit trajectory 
covers regions in the preimage component (observable) that would remain undetectable in conventional analysis.

Our paper is organized as follows: Section \ref{sec:model} presents the \texttt{StarryStarryProcess} model; 
Section \ref{sec:experiment1} presents results on non-evolving stellar surfaces; Section \ref{sec:experiment2} 
presents results on evolving surfaces; Section \ref{sec:toi3884} presents results on TESS data for TOI-3884; and finally,
 we discuss implications and assumptions in Section \ref{sec:discussion}.

\begin{figure}[hbt!]
    \includegraphics[width=0.46\textwidth]{figures/star-planet-cartoon.png}
        \caption{Light curves from the kinds of stars shown on the left panel (these light curves were calculated analytically 
        using \texttt{starry} \citep{Luger2019}): (a) the light curve shows a flat flux if the star has no surface features and no planet orbiting it; 
        (b) a transit in the light curve of the star that hosts an exoplanet; (c) if the star has one orbiting planet and one starspot under the planet's 
        trajectory then it is seen as a bump in a transit signal; (d) the light curve becomes complicated when the star has multiple spots.}
        \label{fig:cartoon}
    \end{figure}

\begin{figure*}[hbt!]
    \includegraphics[width=\textwidth]{figures/nullspace-preimage.png}
        \caption{The breakdown of a surface map into its constituent components: the original map (left), its preimage (center), and null space (right). 
        Accompanying each set is the corresponding impact on the star's rotational light curve. The preimage represents surface patterns that 
        directly influence the observed light curve, while the null space encompasses patterns that have no effect on the flux measurements. 
        It's noteworthy that a significant portion of the surface features fall within the null space, meaning they do not contribute to the light curve 
        that we can observe and measure. The blck line shows the planetary trajectory, so what is important is that the preimage contains the full
        information under planetary trajectory, while the nullspace is missing it.}
        \label{fig:nullspace}
    \end{figure*}

\section{Hierarchical Bayesian model}
\label{sec:model}
In this section, we describe our Gaussian Process (GP) model and the likelihood calculation process used to estimate the model parameters.
This essentially means that we approximate the likelihood 
as a multidimensional Gaussian distribution characterized by a mean $\pmb{\mu}$ and covariance $\pmb{B}$ (for a comprehensive 
derivation of the Gaussian Process, refer to \cite{Luger2021b}):
%
\begin{align}
    \mathcal{L}\left(i_\star, P_\star, \mathbf{u}, \pmb{\theta}_\bullet\right) \sim
    \mathcal{N}\Big(
    \pmb{\mu}\left(i_\star, P_\star, \mathbf{u}, \pmb{\theta}_\bullet\right),
    \,
    \pmb{B}\left(i_\star, P_\star, \mathbf{u}, \pmb{\theta}_\bullet\right)
    \Big).
\end{align}
%
To infer the physical properties of the starposts, we want to obtain an underlying distribution of the physics conditioned on the provided data, 
which is known as the posterior distribution of the model.
% \subsection{Model for non-evolvng surfaces} \label{sec:nonevolmodel} 
We want to solve for a large set of parameters that includes the GP hyperparameters, the star's and the planet's orbital parameters, respectively. 
\begin{linenomath}\begin{align}
    \label{eq:largetheta}
    \pmb{\Theta}
     & =
    \left(
    \theta_\bullet
    \,\,\,
    \theta_\star
    \,\,\,
    \theta_p
    \right)^\top
    \quad,
\end{align}\end{linenomath}

@@

We represent the GP hyperparameters as \emph{physically interesting} set of parameters $\pmb{\theta}_\bullet$ \citep{Luger2021b}:
%
\begin{linenomath}\begin{align}
        \label{eq:thetaspot}
        \pmb{\theta}_\bullet
         & =
        \left(
        \mathbb{n}
        \,\,\,
        \mathbb{c}
        \,\,\,
        \sigma^2_\phi
        \,\,\,
        \mu_\phi
        \,\,\,
        \mathbb{r}
        \right)^\top
        \quad,
    \end{align}\end{linenomath}
%
where $\mathbb{n}$ is the number of starspots, $\mathbb{c}$ is their contrast (defined as the intensity difference between the spot and the 
background intensity, as a fraction of the background intensity), and $\mathbb{r}$ is the radius
of the spots, $\mu_\phi$ and $\sigma^2_\phi$ are the mean and variance of the latitude distribution. 
It could be more convenient to sample $\mu_\phi$ and $\sigma^2_\phi$ using $\mathbb{a}$ and $\mathbb{b}$, where
$\mathbb{a}$ and $\mathbb{b}$ are the normalized parameters of the Beta distribution in $\cos\phi$, which is the probability density function (PDF) 
for the latitude $\phi$ of the spots. These parameters have a one-to-one correspondence to the mode $\mu_\phi$ and standard deviation $\sigma_\phi$ of the 
distribution in $\phi$, allowing for a concise representation of the latitudinal distribution characteristics.
The Beta distribution in $\cos\phi$ which has hyperparameters $\alpha$ and $\beta$, and the PDF given by
%
\begin{align}
    \label{eq:cosphi-pdf}
    p \big(\cos\phi \, \big| \, \alpha, \beta \big)
     & =
    \dfrac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)}
    (\cos\phi)^{\alpha - 1}
    (1 - \cos\phi)^{\beta - 1}
    \quad,
\end{align}
%
where $\Gamma$ is the Gamma function. The $\alpha$ and $\beta$ are derived from the normalized parameters 
%
\begin{align}
    \label{eq:beta2gauss}
    \alpha & = \exp\left({K_{00} + (\ln\frac{1}{2})\mathbb{a}}\right)
    \nonumber                                                 \\
    \beta  & = \exp\left({\ln\frac{1}{2} + (10 - \ln\frac{1}{2})\mathbb{b}}\right)
    \quad,
\end{align}
%
with inverse transform
%
\begin{align}
    \label{eq:gauss2beta}
    \mathbb{a} & \equiv \frac{\ln\alpha}{\ln\frac{1}{2}}
    \nonumber                                             \\[0.5em]
    \mathbb{b} & \equiv \frac{\ln\beta - \ln\frac{1}{2}}{10 -\ln\frac{1}{2}}
\end{align}
%
The parameters $\mathbb{a}$ and $\mathbb{b}$ are both constrained to values between 0 and 1, making them convenient for sampling during the inference process. 
However, $\mathbb{a}$ and $\mathbb{b}$ do not have a straightforward relationship with physically meaningful quantities. In many situations, it is preferable to parametrize 
the latitude distribution using two parameters: $\mu_\phi$, which controls the central latitude, and $\sigma_\phi$, which governs the dispersion of the 
spots' latitudes. Here, both the mean $\mu_\phi$ and variance $\sigma^2_\phi$ can be derived from the Beta distribution as
%
\begin{align}
    \label{eq:mean_beta}
    \mu_\phi
     & =
    \dfrac{\alpha}{\alpha+\beta}
    \quad,
\end{align}
%
\begin{align}
    \label{eq:var_beta}
    \sigma^2_\phi
     & =
    \dfrac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}
    \quad,
\end{align}
%

@@

For faster calculations we rewrite the likelihood function using \emph{the matrix conversion lemma}, also known as 
Woodbury-Sherman-Morrison identity \citep[see, e.g.,][]{Hogg2020}, which gives an expression for the inverse $\pmb{B}^{-1}$ of the marginalized likelihood variance:
%
\begin{linenomath}\begin{align}
    \label{eq:Hoggtrick}
    \pmb{B}^{-1} = \pmb{C}^{-1} - \pmb{C}^{-1} \pmb{M} (\pmb{\mathbb{I}} 
    & + \pmb{\Lambda} \pmb{M}^T \pmb{C}^{-1}\pmb{M})^{-1}\pmb{\Lambda} \pmb{M}^T\pmb{C}^{-1}
\end{align}\end{linenomath}
%

@@

% \subsection{Step-by-step model instruction}
% In this section, we provide a comprehensive, step-by-step description of our model's methodology. 
% This process outlines the key stages in our analysis, from initial data preparation to the final likelihood computation:

% \begin{enumerate}
%     \item Acquire the target system's light curve data.
%     \item Preprocess the data by binning the out-of-transit observations to optimize computational efficiency while preserving signal fidelity.
%     \item Construct a design matrix incorporating both planetary (transit) and stellar orbital parameters, encapsulating the geometry of the star-planet system.
%     \item Establish physically motivated hyperparameters and prior distributions based on current astrophysical understanding and observational constraints.
%     \item Implement a sampling algorithm to draw from the defined prior distributions, exploring the parameter space efficiently.
%     % \item Generate synthetic light curves from the sampled parameters, simulating the observed stellar and planetary signals.
%     \item Compute the covariance matrix for the Gaussian process, which models the stochastic components of stellar variability.
%     \item Assemble the inverse of the full covariance matrix, combining deterministic and stochastic elements of the model based on Equation \ref{eq:Hoggtrick}.
%     \item Calculate the likelihood of the observed data given the model parameters, quantifying the agreement between our model and the observations with Equation \ref{eq:log-likeRodrigo}.
% \end{enumerate}

@@

\section{Results: synthetic datasets}
\label{sec:results-synthetic}
\subsection{Priors}
Before presenting the results, it is essential to elucidate the prior distributions employed throughout this work. 
We have implemented multiple priors in this work, carefully selected to maintain a balance between informativeness and generalizability. 
These priors serve to encapsulate our existing knowledge and assumptions about the parameters of interest, while allowing for sufficient flexibility 
to avoid computational cost.
\begin{itemize}
    \item $\sim\mathcal{U}(\rm low, \rm high)$: a uniform distribution.
    \item $\sim \rm Log \mathcal{U}(\rm low, \rm high)$: a log-uniform distribution.
    \item \textit{Planetary Inclination}: Considering that $i_p=\arccos{b}$, where $b$ is the impact parameter,
    the prior on the inclination is defined via the uniform prior $\sim\mathcal{U}(-b_{max}, -b_{max})$ on $b$, where $b_{max} = R_{\star} / a$.
    \item \textit{Period}: a prior on $P$ and $P_{\star}$. Instead of directly constraining the period, this prior works in log-space to ensure the period 
    remains positive and to make the distribution more symmetric. The actual prior is a uniform distribution in log-space, bounded by the log of the true 
    value plus or minus some fraction. The final period is determined by exponentiating the result of this log-uniform distribution, 
    which effectively creates a kind of log-uniform prior on the period itself.
    \item \textit{Stellar Angle}: a normal distribution on orientation vectors $\tilde{x}$ and $\tilde{y}$ and a half-normal distribution on $\tilde{z}$. See Section
    \ref{sec:obl-inc} for the full explanation.
\end{itemize}

@@

We used No-U-Turn Sampling, a variant of Hamiltonian Monte Carlo \citep[NUTS;][]{Duane1987,Hoffman2011} to do posterior inference on our synthetic dataset.
